---
title: "Analysis"
---

# Global Trends

## Library Preparation



```{r}
#| message: false
#| warning: false
#| include: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(readr)
library(readxl)
library(ggthemes)
library(forcats)
```

## Income Based Mean Emission

Let's first visualize how the mean carbondioxide emissions has changed by countries' income level.

## Income Based Mean Emission

Let's first visualize how the mean carbondioxide emissions has changed by countries' income level.

```{r load-workspace}
#| include: false
load("emu430_project_workspace.RData")
```


```{r, F}
countries_by_income <- c(
  "Upper-middle-income countries", 
  "Lower-middle-income countries", 
  "Low-income countries", 
  "High-income countries"
)

total_co2_emission_by_income <- my_data3 |> 
  filter(entity %in% countries_by_income) |>
  rename(income_class = entity) |>
  group_by(income_class) |>
  summarize(total_co2_income = sum(transport_co2_emissions, na.rm = TRUE))

total_co2_emission_by_income
```

```{r}
ggplot(total_co2_emission_by_income, aes(x = reorder(income_class, total_co2_income), y = total_co2_income, fill = income_class)) +
  geom_col() +
  scale_y_continuous(labels = label_number(scale = 1, accuracy = 1)) +
  coord_flip() +
  labs(x = "Income Class",
       y = "Total CO2 Emission",
       title = "Total CO2 Emission by Income Class",
       subtitle = "From 2011 to 2021 (in tons)") +
  theme_light() +
  theme(legend.position = "none")
```

As expected, the more income levels, the high emission levels that a country would have.

## Country Based Mean Emission

Let's take a final look at the top 20 countries' mean carbondioxide emission.

```{r}
non_countries <- c(
  "World", 
  "Upper-middle-income countries", 
  "Lower-middle-income countries", 
  "Low-income countries", 
  "High-income countries", 
  "European Union (27)", 
  "Europe", 
  "Asia", 
  "Africa",
  "North America",
  "South America",
  "Oceania"
)

mean_co2_emission_by_country <- my_data3 |> 
  filter(!entity %in% non_countries) |>
  group_by(entity) |>
  summarize(mean_co2_emission_per_year = mean(transport_co2_emissions, na.rm = TRUE)) |>
  mutate(entity_lumped = fct_lump_n(entity, n = 20, w = mean_co2_emission_per_year)) |>
  group_by(entity_lumped) |>
  summarize(mean_co2_emission_per_year = mean(mean_co2_emission_per_year, na.rm = TRUE), .groups = "drop")

head(mean_co2_emission_by_country)
```

```{r}
ggplot(mean_co2_emission_by_country, aes(x = reorder(entity_lumped, mean_co2_emission_per_year), y = mean_co2_emission_per_year, fill = entity_lumped)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = label_number(scale = 1, accuracy = 1)) +
  coord_flip() +
  labs(x = "Country",
       y = "Mean CO2 Emission",
       title = "Mean CO2 Emission by Country",
       subtitle = "From 2011 to 2021 (in tons)") +
  theme_light() +
  theme(legend.position = "none")
```

The big majority of mean carbondioxide emissions consisted of top 20 countries.

Scandinavia countries, such as Denmark and Norway, are really worth to analyze their carbondioxide emissions and electric/non-electric EV sales, as they have not listed in the top 20 countries with the largest mean carbondioxide emission level.

Turkey is 19th country with the largest mean carbondioxide emission level.

The amounts of their carbondioxide emission (in tons) are displayed in the graph below.

```{r}
ggplot(emission_by_country, aes(x = year, y = transport_co2_emissions, color = entity)) +
  geom_line(size = 1) +
  scale_y_continuous(
    trans = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 10000000000),
    labels = c("1", "10", "100", "1K", "10K", "100K", "1M", "10M", "100M", "1B", "10B")
  ) +
  scale_x_continuous(breaks = 2011:2021) +
  facet_wrap(~ entity) +
  labs(x = "Year", 
       y = "CO2 Emission",
       title = "Carbondioxide Emissions by Entity (Logarithmic Scale)",
       subtitle = "From 2011 to 2021 (in tons)"
       ) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1)) # rotate x-axis labels

emission_by_country
```

1.  It can be observed that there is a persistent increase in the carbondioxide emission levels in Turkey and China. China is very close to emit one billion tons of carbondioxide.

2.  Denmark has decreased its carbondioxide emission level after a 10-year horizon. Japan and Norway has a considerable decreasing trend in its carbondioxide emission level.

3.  USA has the highest carbondioxide emission compared to others, exceeding one billion tons of carbondioxide.

The line plot below displays the trend in the carbondioxide emission levels of Turkey and including the ones that are on the top of the bar plot of mean carbondioxide emission level, such as USA and China, along with an another country that should be considered, Japan.

```{r}
top6_countries <- c("United States", "Norway", "Denmark", "China", "Japan", "Turkey")

emission_by_country <- my_data3 |>
  filter(entity %in% top6_countries)

head(emission_by_country)
```

```{r}
ggplot(emission_by_country, aes(x = year, y = transport_co2_emissions, color = entity)) +
  geom_line(size = 1) +
  scale_y_continuous(
    trans = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 10000000000),
    labels = c("1", "10", "100", "1K", "10K", "100K", "1M", "10M", "100M", "1B", "10B")
  ) +
  scale_x_continuous(breaks = 2011:2021) +
  facet_wrap(~ entity, scale = "free_y") +
  labs(x = "Year", 
       y = "CO2 Emission",
       title = "Carbondioxide Emissions by Entity (Logarithmic Scale)",
       subtitle = "Carbondioxide emission trends from 2011 to 2021"
       ) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1)) # rotate x-axis labels
```

COVID-19 pandemic has affected the carbondioxide emission levels in a positive way, resulting in kind of a V-shaped pattern for all countries listed.

# Global EV Sales

What about the other countries' adoption on EV's?

```{r}
top4_countries <- c("United States", "Denmark", "China", "Japan")

sales_by_country <- my_data7 |>
  filter(entity %in% top4_countries, year %in% 2011:2021) |>
  mutate(electric_sales_perc = electric_car_sales / sum(electric_car_sales, non_electric_car_sales, na.rm = TRUE),
         non_electric_sales_perc = non_electric_car_sales / sum(electric_car_sales, non_electric_car_sales, na.rm = TRUE)) |>
  pivot_longer(cols = c(electric_sales_perc, non_electric_sales_perc),
               names_to = "car_sales",
               values_to = "value"
  )

head(sales_by_country)
```

```{r}
ggplot(sales_by_country, aes(x = year, y = value, fill = car_sales)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(breaks = 2011:2021) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "Year",
       y = "Sales",
       title = "Car Sales by Country",
       subtitle = "Proportion of EV's and Non-EV's"
  ) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + # rotate x-axis labels
  facet_wrap(~ entity)
```

Note that these sales are just cars sales.

The types of EV's are summarized below.

-   Battery Electric Vehicle (BEV)
-   Plug-in Hybrid Electric Vehicle (PHEV)
-   Fuel Cell Electric Vehicle (FCEV)
-   Hybrid Electric Vehicle (HEV)
-   Mild Hybrid Vehicle (MHEV)

The electric cars include Battery Electric Vehicles (BEV's) and Plug-in Hybrid Electric Vehicles (PHEV's) for this plot.

It can be observed that Denmark has adopted EV's faster than other countries in the plot.

```{r}
my_data2$region[my_data2$region == "Turkiye"] <- "Turkey"
top6_countries = c("USA", "Norway", "Denmark", "China", "Japan", "Turkey")
no_carbon = c("BEV", "FCEV")

top6_ev_sales <- my_data2 |>
  filter(region %in% top6_countries, parameter == "EV sales", powertrain %in% no_carbon)

top6_ev_sales |> mutate(value = as.numeric(format(value, scientific = FALSE)))

head(top6_ev_sales)
```

```{r}
top6_ev_sales |> group_by(year, region, powertrain) |>
  summarize(total_sales = sum(value, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = total_sales, color = region, linetype = powertrain)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = 2011:2021) +
  scale_y_continuous(trans = "log10",
                     labels = label_number(scale = 1, accuracy = 1)
                     ) +
  labs(x = "Year",
       y = "Sales",
       title = "Sales of No Carbon Vehicles (Logarithmic Scale)",
       subtitle = "The behaviour of the number of sales of vehicles that emit no carbon over time",
       color = "Region",
       linetype = "Powertrain Type"
  ) +
  facet_wrap(~ region) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + # rotate x-axis labels
  guides(color = "none") # remove the region legend
  
top6_ev_sales
```

All countries in the plot increased their non-carbon emitting vehicle sales after 10-year horizon. However, Turkey has not been adopted FCEV's through 2021 while the other countries are using FCEV's more year by year. China has started using FCEV's (in 2016) later than the others (in 2011-2014). It can be also observed that BEV sales has started 2012 in Turkey.

# Turkey's Status

## Vehicle Trends

How about the adoption of EV vehicles and the others in Turkey?

```{r}
tr_vehicle_num <- my_data4 |>
  mutate(across(everything(), as.numeric)) |>
  select(-starts_with("percentage"))

tr_vehicle_num_long <- tr_vehicle_num |>
  pivot_longer(
    cols = c(gas, diesel, lpg, hybrid, electric, unknown), 
    names_to = "vehicle_type", 
    values_to = "value"        
  )

head(tr_vehicle_num_long)
```

These are provided in the line plots below.

```{r}
ggplot(tr_vehicle_num_long, aes(x = year, y = value, color = vehicle_type)) +
  geom_line(size = 1) +
  labs(x = "Year",
       y = "Value",
       title = "Vehicle Trends in Turkey Over Time (Logarithmic Scale)",
       subtitle = "From 2011 to 2021"
  ) +
  scale_x_continuous(breaks = 2011:2021) +
  scale_y_continuous(
    trans = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000),
    labels = c("1", "10", "100", "1K", "10K", "100K", "1M", "10M")
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)
  )
```


Although there was an increase in EV's and hybrid vehicles, Turkey's carbondioxide emission has increased consistently and the effect of COVID-19 pandemic was quite temporary on it.

The cumulative bar plot below displays the proportion of vehicle types over time.

```{r}
vehicle_prop <- c("percentage_gas", "percentage_diesel", "percentage_lpg", "percentage_hybrid", "percentage_electric", "percentage_unknown")

tr_vehicle_perc_long <- my_data4 |>
  pivot_longer(cols = vehicle_prop, names_to = "vehicle_type", values_to = "percentage") |>
  select(year, vehicle_type, starts_with("percentage"))

tr_vehicle_perc_long$percentage <- as.numeric(tr_vehicle_perc_long$percentage)

head(tr_vehicle_perc_long)
```

```{r}
ggplot(tr_vehicle_perc_long, aes(x = year, y = percentage, fill = vehicle_type)) +
  geom_bar(stat = "identity", position = "fill", width = 0.8) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "Year",
       y = "Percentage",
       title = "Vehicle Propensity in Turkey Over Time",
       subtitle = "From 2011 to 2021"
  ) +
  theme_tufte()
```

In fact, non-EV's were still used widely in Turkey, leading to increasing carbondioxide emission levels.

## Mean Emission

Turkey's average carbondioxide emission per vehicle is displayed in the line plot below.

```{r}
tr_emission <- my_data3 |>
  filter(entity == "Turkey")

tr_vehicle <- my_data4 |>
  select(year, total)

tr_vehicle$year <- as.numeric(tr_vehicle$year)

tr_emission_n_vehicle <- tr_emission |>
  left_join(tr_vehicle, by = c("year"))

tr_emission_n_vehicle$total <- as.numeric(tr_emission_n_vehicle$total)
tr_emission_n_vehicle$year <- as.numeric(tr_emission_n_vehicle$year)

head(tr_emission_n_vehicle)
```

```{r}
tr_avg_emission <- tr_emission_n_vehicle |>
  mutate(avg_co2_emission_per_vehicle = transport_co2_emissions / total) |>
  ggplot(aes(x = year, y = avg_co2_emission_per_vehicle)) +
  geom_line(color = "blue", size = 1) +
  scale_x_continuous(breaks = 2011:2021) +
  geom_vline(xintercept = 2020, color = "black", size = 0.5, linetype = "dashed") +
  geom_text(x = 2020, y = 5.5, label = "pandemic", color = "black", vjust = -1) +
  labs(x = "Year",
       y = "Average CO2 Emission per Year",
       title = "Average CO2 Emission Per Vehicle in Turkey (in tons)",
       subtitle = "How much carbondioxide does a vehicle emit on the average?"
  ) +
  theme_minimal()

tr_avg_emission
```