---
title: "Data"
---

# Data Source

1)[my_data2](https://www.kaggle.com/datasets/fathyfathysahlool/electric-vehicles-cars-2011-2024)

2)[my_data3](https://ourworldindata.org/grapher/co2-emissions-transport?time=latest)

3)[my_data4](https://data.tuik.gov.tr/Bulten/Index?p=Motorlu-Kara-Tasitlari-Ekim-2024-53461)

```{r load-workspace}
#| include: false
load("emu430_project_workspace.RData")
```

# GLOBAL EV SALES DATA (my_data2)

This data is used for detecting global electric vehicle sales by country and across the world.

To download this data, the following codes are used.

```{r}
#| warning: false
library(readr)
library(readxl)

# my_data2 <- read_csv("C:/Users/cerem/downloads/IEA Global EV Data 2024.csv")

# head(my_data2, 10)
```

# CO2 EMISSION FROM TRANSPORT DATA (my_data3)

This data is used to determine vehicle-related CO??? emissions globally and by country.

To download this data, the following codes are used.

```{r}
library(jsonlite)

# Fetch the data
my_data3 <- read.csv("https://ourworldindata.org/grapher/co2-emissions-transport.csv?v=1&csvType=full&useColumnShortNames=true")

# Fetch the metadata
metadata <- fromJSON("https://ourworldindata.org/grapher/co2-emissions-transport.metadata.json?v=1&csvType=full&useColumnShortNames=true")

head(my_data3, 10)
```

# TURKIYE'S VEHICLE DATA (my_data4)

This data is used to analyse Turkiye's status about EVs.

To download this data, the following codes are used.

```{r}
library(readxl)

#The following code is used to download the data from excel. Necessary adaptions must be made according to your localcomputer. 

#my_data4 <- read_excel("C:/dosyalar/veriler.xls",sheet = 1)

# head(my_data4, 10)
```
